---
title: "Flight Delays"
params:
  dow: "Saturday"
bibliography: delays.bib
output: ioslides_presentation
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(nycflights13) # data
library(dplyr)        # cleaning functions
library(ggplot2)      # plotting functions
library(lubridate)    # date-time processing

dow <- params$dow

delays <-
  flights %>%
  mutate(date = ymd(paste(year, month, day)), 
         weekday = wday(date, label = TRUE, abbr = FALSE)) %>% 
  select(origin, date, weekday, dep_delay) %>%
  filter(!is.na(dep_delay))

year <-
  delays %>% 
  group_by(origin, date) %>% 
  summarise(mean_delay = mean(dep_delay))

weekday <- 
  delays %>%
  filter(weekday == dow) %>%
  group_by(origin) %>%
  summarise(mean_delay = mean(dep_delay))
```

## Goal

Which airport near New York City has the shortest departure delays on `r dow`?

## Motivation

Airline passengers have choices when they choose to fly in to or out of New York City. Three major airports serve the area:

* **JFK** - John F Kennedy International Airport
* **LGA** - LaGuardia Airport
* **EWR** - Newark Liberty International Airport

How should passengers choose which airport to depart from? This study compares the airports on an obvious criteria: _departure delays_, the amount of time that passes between when a plane is scheduled to depart and when it actually departs. The amount of delay varies by ariport and fluctuates throughout the year, as seen in the graph below.

```{r echo = FALSE}
ggplot(year, aes(x = date, y = mean_delay, color = origin)) + 
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  ggtitle("Smoothed daily mean delays") + 
  ylab("Mean delay (m)") + 
  theme_bw()
```


Since passengers can normally plan which day of the week they travel on, we will condition our analysis on a single day of the week: `r dow`.

## Data

We will examine all flights that originated from a New York City airport in 2013. The data is provided by the Bureau of Transportation Statistics (BTS) and is available for download from <http://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236>. We access the data through the `nycflights13` R package [@wickham14].

## Conclusion

```{r echo = FALSE}
ggplot(weekday, aes(x = origin)) + 
  geom_bar(aes(weight = mean_delay)) +
  ggtitle(paste("Expected", dow, "departure delay", sep = " ")) + 
  ylab("Mean delay (m)") +
  xlab("")
```

As the plot above shows, `r c("EWR" = "Newark", "JFK" = "JFK", "LGA" = "LaGuardia")[[weekday$origin[which.min(weekday$mean_delay)]]]` Airport had the shortest mean departure delay on `r dow`, with a mean delay of `r round(min(weekday$mean_delay), 2)` minutes.

The complete results are in the table below.

```{r echo = FALSE}
knitr::kable(weekday)
```

## References
